Safe Transferable Regions

### 2016-11-29 paper/abstract.tex
Re-added allocation context to type judgment. Restores completeness of type inference.

Manual memory management is susceptible to the usual perils of
dangling pointers. A key contribution of this paper is a
refinement-based region type system, for a C\#-like typed object-oriented
language with higher-order functions and generics, that ensures the
memory safety of programs in the presence of transferable regions.
We complement our type system with a type inference algorithm, which eliminates the need for programmers to write region annotations on types.
% The cornerstone of our inference algorithm is a novel constraint solver that
% performs abduction in a partial-order constraint domain to infer weakest solutions
% to recursive constraints.
We establish soundness and partial completeness results for the type inference algorithm: in particular, the key steps of constraint-generation and constraint-solving in the type inference algorithm are complete. We describe a prototype implementation of our inference algorithm and the constraint solver, and our experience of using it to enforce memory safety in dataflow programs.

>>>>

Manual memory management is susceptible to the usual perils of
dangling pointers. A key contribution of this paper is a
refinement-based region type system, for a C\#-like typed
object-oriented language with higher-order functions and generics,
that ensures the memory safety of programs in the presence of
transferable regions.  We complement our type system with a type
inference algorithm, which eliminates the need for programmers to
write region annotations on types.
% The cornerstone of our inference algorithm is a novel constraint solver that
% performs abduction in a partial-order constraint domain to infer weakest solutions
% to recursive constraints.
The inference algorithm has been proven sound, and the constraint
solving algorithm used by the inference has been proven complete.  We
describe a prototype implementation of the inference algorithm and the
solver, and our experience of using it to enforce memory safety in
dataflow programs.

### 2016-11-29 paper/introduction.tex
Removed claims of completeness.
Re-added allocation context to type judgment. Restores completeness of type inference.

Our inference algorithm proceeds in three stages: in the first stage,
we elaborate the program by introducing region parameters (for classes and methods);
in the second stage we generate a set of constraints that must be satisfied for
the program to type check; in the third stage, we solve the set of constraints,
inferring the preconditions of all classes and methods (in terms of the
expected outlives-constraints between the region parameters).
We show that the second and third stages are sound as well as
complete (while the first stage is sound).
% (using an interprocedural fixed point computation).

@@

\item We establish partial completeness results for the type inference
  algorithm, as well as soundness.

>>>>

Our inference algorithm proceeds in three stages: in the first stage,
we elaborate the program by introducing region parameters (for classes
and methods); in the second stage we generate a set of constraints
that must be satisfied for the program to type check; in the third
stage, we solve the set of constraints, inferring the preconditions of
all classes and methods (in terms of the expected outlives-constraints
between the region parameters).  We show that the algorithm is sound,
and the last stage of constraint solving is complete.
% (using an interprocedural fixed point computation).

@@

\item We establish soundness of type inference algorithm, and
completeness of constraint solving.

### 2016-11-29 paper/type-inference.tex
Re-added allocation context to type judgment. Restores completeness of type inference.

We later on establish the soundness of the type inference algorithm, and also establish
that the constraint-generation and constraint-solving steps (the second and third steps
above) are complete.

@@

The set $\satC$ also makes it easy to check if $C$ is satisfiable.
Let $\myground{C}$ denote the subset of all ground constraints
(\ie, constraints without any region variable or predicate variable)
in $\saturate{C}$. Let $\rhoC$ denote the subset of all wellformedness
constraints for region variables in $\satC$.
Define $\solveCon{C}$ as below.

>>>>

We later on establish the soundness of the type inference algorithm,
and the completeness of constraint-solving step.

@@

The set $\satC$ also makes it easy to check if $C$ is satisfiable.
Let $\myground{C}$ denote the subset of all ground constraints
(\ie, constraints without any region variable or predicate variable)
in $\saturate{C}$. Let $\rhoC$ denote the subset of all well-formedness
constraints for region variables in $\satC$.
Define $\solveCon{C}$ as below.

### 2016-11-29 paper/type-system.tex
Re-added allocation context to type judgment. Restores completeness of type inference.

Fig~\ref{fig:fb-syntax} describes the syntax of \FB.
We refer to the class types of FGJ as \emph{core types}.
%
The following definition of \C{Pair} class in \FB illustrates some of
the key elements of the formal language (the symbol $\extends$ should
be read \emph{extends}, and the symbol $\outlives$ stands for
\emph{outlives}):
\begin{codejava}[mathescape=true]
class Pair<a $\extends$ Object, b $\extends$ Object>
          <$\rho^a, \rho_1, \rho_2 \,|\, \rho_1 \outlives \rho^a \conj \rho_2 \outlives \rho^a$> $\extends$ Object<$\rho^a$> {
  a@$\rho_1$ fst; 
  b@$\rho_2$ snd;
  a@$\rho_1$ getFst() { return this.fst; }
}
\end{codejava}
% \vspace*{-0.15in}
%  Pair(a@$\rho_1$ fst, b@$\rho_2$ snd) {
%    super(); this.fst = fst; this.snd = snd;
%  }
Note that we elide showing constructors since they are uninteresting
from the type system's standpoint; their behavior in \FB is same as in
FGJ.

A class in \FB is parametric over zero or more type variables (as
in FGJ) as well as one or more region variables $\rho$.
% In the above example, the class \C{Pair} is parameterized over
% $\rho^a, \rho_1, \rho_2$.
We refer to the first region parameter, usually denoted $\rho^a$, as
the \emph{allocation region} of the class: it serves to identify the
region where an instance of the class is allocated.
%
An object in \FB can contain fields referring to objects allocated in regions ($\rhobar$) other than its
own allocation region ($\rhoalloc$), provided that the former outlive
the latter (i.e., $\rhobar \outlives \rhoalloc$). In such case, the
definition of object's class needs to be parametric over allocation
regions of its fields (i.e., their classes). Furthermore, the
constraint that such regions must outlive the allocation region of the
class needs to be made explicit in the definition, as the \C{Pair}
class does in the above definition. We say that the \C{Pair} class
exhibits \emph{constrained region polymorphism}.
%
% For instance, a \C{Pair} object allocated in a region ${\rhoalloc}$ can have its first
% component coming from ${\rho_1}$ and second from $\rho_2$, as long as
% $\rho_1, \rho_2$ outlive $\rhoalloc$, as illustrated by the above definition.

To construct objects of the \C{Pair} class, its type and region
parameters need to be instantiated with core types ($T$) and region
annotations\footnote{Region annotations ($\rgn$) include region
variables ($\rho$) and region identifiers ($\pi$). Region identifiers
are to region variables, as types ($T$) are to type variables ($a$)}
($\rgn$), respectively. For example:
\begin{codejava}
letregion $\pi_0$ in
  let snd = new Object<$\pi_0$>() in
  letregion $\pi_1$ in
    let fst = new Object<$\pi_1$>() in
    let p = new Pair<Object,Object><$\pi_1$,$\pi_1$,$\pi_0$> (fst,snd);
\end{codejava}
%\vspace*{-0.1in}
In the above code, the instantiation of $\rhoalloc$ and $\rho_1$ with
$\pi_1$, and $\rho_2$ with $\pi_0$ is allowed because (a) $\pi_0$
and $\pi_1$ are live during the instantiation, and (b). $\pi_0
\outlives \pi_1$ and $\pi_1 \outlives \pi_1$ (since outlives is
reflexive). Observe that the region type of \C{p} conveys
the fact that (a). it is allocated in region $\pi_1$, and (b). it
holds references to objects allocated in region $\pi_0$ and $\pi_1$.
In contrast, if we choose to allocate the \C{snd} object also in
$\pi_1$, then \C{p} would be contained in $\pi_1$, and its region
type would be \C{Pair<\ObjZ,\ObjZ><$\pi_1$,$\pi_1$,$\pi_1$>}, which
we abbreviate as \C{Pair<\ObjZ,\ObjZ>@$\pi_1$}. In general, we treat
$B\inang{\tbar}@\pi$ as being equivalent to
$B\inang{\tbar}\inang{\rbar}$. Region annotation on type $a$, where
$a$ is a type variable, assumes the form $a@\pi$. If $a$ is
instantiated with \C{Pair<\ObjZ,\ObjZ>}, the result is the type of a
\C{Pair} object contained in  $\pi$. Class fields are not allowed to
be type polymorphic, in keeping with C\#, which constitutes the core
of \name. Separate parameterization over types and regions is needed
in \FB to support region-polymorphic higher-order functions as class
fields. For such higher-order class fields, regions are parameterized
at the (class) field level whereas types are parameterized at the
class level. This allows, for example, a generic class, whose type
parameters are $a$ and $b$, to contain a region-polymorphic function of type
$\inang{\rho_0,\rho_1,\rho_2 \,|\, \rho_1 \outlives \rho_0 \conj
\rho_2 \outlives \rho_0}(a@\rho_1,b@\rho_2) \rightarrow
\C{Pair}\inang{a,b}\inang{\rho_0,\rho_1,\rho_2}$ as its field. Such
region-polymorphic higher-order fields are used frequently by the
dataflow operators, which apply them in the context of various regions
(\emph{e.g.,} see Fig.~\ref{fig:motivating-eg-in-broom}). Keeping type
parameterization seperate from region parameterization allows \FB to
type such examples. This choice also simplifies type inference in the
presence of higher-order functions.

@@

Well-formedness and typing rules of \fbname establish the conditions
under which a region type is considered well-formed, and an expression
is considered to have a certain region type, respectively.
Fig.~\ref{fig:fb-staticsem} contains an illustrative subset of such
rules\footnote{Full formal development can be found in the
appendix}. The rules refer to a context ($\A$), which is a tuple of:
\begin{itemize}
\item A set ($\rhoenv \in 2^\rgn$) of regions that are estimated to be
live,
\item A finite map ($\aenv \in \tyvar \mapsto \fgjN$) of type
variables to their \emph{bounds}, i.e., classes they are declared to
extend (this artifact is inherited from FGJ), and
\item A constraint formula ($\phicx$) that captures the outlives
constraints on regions in $\rhoenv$.
\end{itemize}
In addition, the context for the expression typing judgment also
includes a type environment ($\env \in x \mapsto \tau$) that contains
the type bindings for variables in scope. Like the judgments in
FGJ~\cite{fgj}, all the judgments defined by the rules in
Fig.~\ref{fig:fb-staticsem} are implicitly parameterized on a class
table ($CT \in cn \mapsto D$) that maps class names to their
definitions in \FB.

@@

The type rule for \C{letregion} expression requires that the static
identifier introduced by the expression be unique under the current
context (i.e., $\rgn \notin \rhoenv$). This condition is needed in
order to prevent the new region from incorrectly assuming existing
outlives relationships on an eponymous region. Provided this is
satisfied, the expression ($e$) under \C{letregion} is then
typechecked assuming that the new region is live ($\rgn \in \rhoenv$)
and that it is outlived by all existing live regions ($\rhoenv
\outlives \rgn$). The result of a \C{letregion} expression must have a
type that is well-formed under a context not containing the new
region. This ensures that the value obtained by evaluating a
\C{letregion} expression contains no references to the temporary
objects inside the region.

@@

The type rule for lambda expression typechecks
% requires that its region
% parameters ($\rhoalloc\rhobar$) be unique under the current context,
% and that the constraints ($\phi$) over region parameters be
% well-formed. 
the lambda-bound expression ($e$) under an extended type environment
containing bindings for function's arguments, assuming that region
parameters are live, and that declared constraints over region
parameters hold. The constraints ($\phi$) are required to be
well-formed under $\rhoenv$ extended with the function's region
parameters ($\rhobar$). The premise $\forall
(\rgn'\in\frv(e)\setminus\{\rhobar\}). ~\isvalid{\phicx}
{\rgn'\outlives\rgn}$ ensures the safety of dereferences inside the
closure by requring that all free region variables ($\frv$) trapped by
the closure outlive the allocation region of the closure.

@@

Fig.~\ref{fig:fb-opsem-1} and~\ref{fig:fb-opsem-2} in the appendix
defines a small-step operational semantics for \fbname via a
five-place reduction relation:
% \vspace*{-0.15in}
\begin{smathpar}
  \redstoo{\rhoenv}{(e,\mem)}{(e',\mem')}
\end{smathpar}
Operational semantics replaces abstract region annotations ($\pi$ and
$\rho$) with \emph{locations} ($\loc$), where each \emph{location}
abstracts all the memory locations associated with a region (i.e.,
each memory region is associated with a single location). The
typestate of regions is tracked by a finite map ($\mem$) from
locations to typestates (Fig.~\ref{fig:region-fsm}). The reduction
relation should be read as following: given a set ($\rhoenv$) of
regions that are currently live, and an initial typestate map
($\mem$), the expression $e$ reduces to $e'$, while updating $\mem$ to
$\mem'$. The semantics gets ``stuck'' if $e$ attempts to access an
object whose allocation region is not present in $\rhoenv$, or if $e$
tries to \C{open} a transferable region that is not mapped to an
appropriate typestate by $\mem$.  On the other hand, if $e$ attempts
to commit an operation on a $\RgnZ$ object that is not sanctioned by
the transition discipline in Fig.~\ref{fig:region-fsm}, then it raises
an exception value ($\invalidexn$).

>>>>

Fig~\ref{fig:fb-syntax} describes the syntax of \FB.
We refer to the class types of FGJ as \emph{core types}.
%
The following definition of \C{Pair} class in \FB illustrates some of
the key elements of the formal language (the symbol $\extends$ should
be read \emph{extends}, and the symbol $\outlives$ stands for
\emph{outlives}):
\begin{codejava}[mathescape=true]
class Pair<a $\extends$ Object, b $\extends$ Object>
          <$\rho_0, \rho_1, \rho_2 \,|\, \rho_1 \outlives \rho_0 \conj
          \rho_2 \outlives \rho_0$> $\extends$ Object<$\rho_0$> {
  a@$\rho_1$ fst; 
  b@$\rho_2$ snd;
  a@$\rho_1$ getFst() { return this.fst; }
}
\end{codejava}
% \vspace*{-0.15in}
%  Pair(a@$\rho_1$ fst, b@$\rho_2$ snd) {
%    super(); this.fst = fst; this.snd = snd;
%  }
Note that we elide showing constructors since they are uninteresting
from the type system's standpoint; their behavior in \FB is same as in
FGJ.

A class in \FB is parametric over zero or more type variables (as
in FGJ) as well as one or more region variables $\rho$.
% In the above example, the class \C{Pair} is parameterized over
% $\rho^a, \rho_1, \rho_2$.
We refer to the first region parameter ($\rho_0$ in the above example)
as the \emph{allocation region} of the class: it serves to identify
the region where an instance of the class is allocated\footnote{In
general, $\rhobar$ denotes the sequence $\rho_1\rho_2...$ of region
parameters of a class or a method. In some cases, we also represent
region parameters as $\rho\rhobar$ to clearly distinguish between
the allocation region parameter ($\rho$) and the rest.}.
%
An object in \FB can contain fields referring to objects allocated in
regions ($\rhobar$) other than its own allocation region ($\rho$),
provided that the former outlive the latter (i.e., $\rhobar \outlives
\rho$). In such case, the definition of object's class needs to be
parametric over allocation regions of its fields (i.e., their
classes). Furthermore, the constraint that such regions must outlive
the allocation region of the class needs to be made explicit in the
definition, as the \C{Pair} class does in the above definition. We say
that the \C{Pair} class exhibits \emph{constrained region
polymorphism}.
%
% For instance, a \C{Pair} object allocated in a region ${\rhoalloc}$ can have its first
% component coming from ${\rho_1}$ and second from $\rho_2$, as long as
% $\rho_1, \rho_2$ outlive $\rhoalloc$, as illustrated by the above definition.

To construct objects of the \C{Pair} class, its type and region
parameters need to be instantiated with core types ($T$) and region
annotations\footnote{Region annotations ($\rgn$) include region
variables ($\rho$) and region identifiers ($\pi$). Region identifiers
are to region variables, as types ($T$) are to type variables ($a$)}
($\rgn$), respectively. For example:
\begin{codejava}

letregion $\pi_0$ in
  let snd = new Object<$\pi_0$>() in
  letregion $\pi_1$ in
    let fst = new Object<$\pi_1$>() in
    let p = new Pair<Object,Object><$\pi_1$,$\pi_1$,$\pi_0$> (fst,snd);
\end{codejava}
%\vspace*{-0.1in}
In the above code, the instantiation of $\rho_0$ and $\rho_1$ with
$\pi_1$, and $\rho_2$ with $\pi_0$ is allowed because (a) $\pi_0$
and $\pi_1$ are live during the instantiation, and (b). $\pi_0
\outlives \pi_1$ and $\pi_1 \outlives \pi_1$ (since outlives is
reflexive). Observe that the region type of \C{p} conveys
the fact that (a). it is allocated in region $\pi_1$, and (b). it
holds references to objects allocated in region $\pi_0$ and $\pi_1$.
In contrast, if we choose to allocate the \C{snd} object also in
$\pi_1$, then \C{p} would be contained in $\pi_1$, and its region
type would be \C{Pair<\ObjZ,\ObjZ><$\pi_1$,$\pi_1$,$\pi_1$>}, which
we abbreviate as \C{Pair<\ObjZ,\ObjZ>@$\pi_1$}. In general, we treat
$B\inang{\tbar}@\pi$ as being equivalent to
$B\inang{\tbar}\inang{\rbar}$. Region annotation on type $a$, where
$a$ is a type variable, assumes the form $a@\pi$. If $a$ is
instantiated with \C{Pair<\ObjZ,\ObjZ>}, the result is the type of a
\C{Pair} object contained in  $\pi$. Class fields are not allowed to
be type polymorphic, in keeping with C\#, which constitutes the core
of \name. Separate parameterization over types and regions is needed
in \FB to support region-polymorphic higher-order functions as class
fields. For such higher-order class fields, regions are parameterized
at the (class) field level whereas types are parameterized at the
class level. This allows, for example, a generic class, whose type
parameters are $a$ and $b$, to contain a region-polymorphic function of type
$\inang{\rho_0,\rho_1,\rho_2 \,|\, \rho_1 \outlives \rho_0 \conj
\rho_2 \outlives \rho_0}(a@\rho_1,b@\rho_2) \rightarrow
\C{Pair}\inang{a,b}\inang{\rho_0,\rho_1,\rho_2}$ as its field. Such
region-polymorphic higher-order fields are used frequently by the
dataflow operators, which apply them in the context of various regions
(\emph{e.g.,} see Fig.~\ref{fig:motivating-eg-in-broom}). Keeping type
parameterization separate from region parameterization allows \FB to
type such examples. This choice also simplifies type inference in the
presence of higher-order functions.

@@

Well-formedness and typing rules of \fbname establish the conditions
under which a region type is considered well-formed, and an expression
is considered to have a certain region type, respectively.
Fig.~\ref{fig:fb-staticsem} contains an illustrative subset of such
rules\footnote{Full formal development can be found in the
appendix}. The rules refer to a context ($\A$), which is a tuple of:
\begin{itemize}
\item A set ($\rhoenv \in 2^\rgn$) of regions that are estimated to be
live,
\item A finite map ($\aenv \in \tyvar \mapsto \fgjN$) of type
variables to their \emph{bounds}, i.e., classes they are declared to
extend (this artifact is inherited from FGJ), and
\item A constraint formula ($\phicx$) that captures the outlives
constraints on regions in $\rhoenv$.
\end{itemize}
In addition, the context for the expression typing judgment also
includes (a). a type environment ($\env \in x \mapsto \tau$) that
contains the type bindings for variables in scope, and (b). The region
($\rgn$) that serves as the allocation context for the expression
being type checked. Like the judgments in FGJ~\cite{fgj}, all the
judgments defined by the rules in Fig.~\ref{fig:fb-staticsem} are
implicitly parameterized on a class table ($CT \in cn \mapsto D$) that
maps class names to their definitions in \FB.

@@

The type rule for \C{letregion} expression requires that the static
identifier introduced by the expression be unique under the current
context (i.e., $\pi \notin \Delta$). This condition is needed in order
to prevent the new region from incorrectly assuming existing outlives
relationships on an eponymous region. Provided this is satisfied, the
expression ($e$) under \C{letregion} is then typechecked assuming that
the new region is live ($\Delta \cup \{\pi\}$) and that it is outlived
by all existing live regions ($\Delta \outlives \pi$). The result of a
\C{letregion} expression must have a type that is well-formed under a
context not containing the new region. This ensures that the value
obtained by evaluating a \C{letregion} expression contains no
references to the temporary objects inside the region.

@@

The type rule for lambda expression typechecks the lambda-bound
expression ($e$) under an extended type environment containing
bindings for function's arguments, assuming that region parameters are
live, and that declared constraints over region parameters hold. The
constraints ($\phi$) are required to be well-formed under $\Delta$
extended with the function's region parameters ($\rho\rhobar$). Note
that the closure is always allocated in the current allocation context
($\rgn$). This prevents the closure from escaping the context in which
it is created, thus trivially ensuring the safety of any dereferences
inside the closure.

@@

Fig.~\ref{fig:fb-opsem-1} and~\ref{fig:fb-opsem-2} in the appendix
defines a small-step operational semantics for \fbname via a
four-place reduction relation:
% \vspace*{-0.15in}
\begin{smathpar}
  \redstoo{\rhoenv}{(e,\mem)}{(e',\mem')}
\end{smathpar}
Operational semantics replaces abstract region annotations ($\pi$ and
$\rho$) with \emph{locations} ($\loc$), where each \emph{location}
abstracts all the memory locations associated with a region (i.e.,
each memory region is associated with a single location). The
typestate of regions is tracked by a finite map ($\mem$) from
locations to typestates (Fig.~\ref{fig:region-fsm}). The reduction
relation relates an expression ($e$) and a typestate map ($\mem$) to a
reduced expression ($e'$) and an updated typestate map ($\mem'$). The
semantics gets ``stuck'' if $e$ attempts to access an object whose
allocation region is not present in $\rhoenv$, or if $e$ tries to
\C{open} a transferable region that is not mapped to an appropriate
typestate by $\mem$.  On the other hand, if $e$ attempts to commit an
operation on a $\RgnZ$ object that is not sanctioned by the transition
discipline in Fig.~\ref{fig:region-fsm}, then it raises an exception
value ($\invalidexn$).

### 2016-11-29 paper/typeinference-proofs.tex
minor fixes
Removed claims of completeness.
Re-added allocation context to type judgment. Restores completeness of type inference.

\newcommand{\deltaPC}{\Delta_P^C}

@@

@@

\begin{proof}[\textbf{Theorem~\ref{thm:constraint-generation-sc}}]
The correspondence between the static semantics rules and the constraint generation
rules induces a correspondence between derivation trees produced by the
two systems.
% We do not present a formal proof of the theorem.
We can establish the following lemma inductively:
given a derivation tree $\zeta_1$ for
$\exprok {\stdcontext}{e}{\tau}{C}$
and any substitution $\soln$,
we can construct a corresponding \emph{candidate} derivation
tree $\zeta_2$ of $\hastyp {\A[\soln],\env[\soln]}{e[\soln]}{\tau[\soln]}$.
We can show that $\zeta_2$ is a valid derivation tree iff $\soln$ satisfies $C$.
(We make use of analogous lemmas for type well-formedness rules as well.)
\qed
\end{proof}

>>>>

@@

Figs.~\ref{fig:constraint-gen-1} and~\ref{fig:constraint-gen-2} contain the
extra constraint-generation rules not covered in Fig.~\ref{fig:constraint-gen-0}.
We now present a sketch of the proofs of the soundness of type
inference algorithm, and completeness of the constraint solver.

@@

\begin{proof}[\textbf{Theorem~\ref{thm:constraint-generation-sc}}]
The correspondence between the static semantics rules and the constraint generation
rules induces a correspondence between derivation trees produced by the
two systems.
% We do not present a formal proof of the theorem.
We can establish the following lemma inductively:
given a derivation tree $\zeta_1$ for
$\exprok {\stdcontext}{e}{\tau}{C}$
and any substitution $\soln$,
we can construct a corresponding \emph{candidate} derivation
tree $\zeta_2$ of $\hastyp {\A[\soln],\env[\soln]}{e[\soln]}{\tau[\soln]}$.
We can show that $\zeta_2$ is a valid derivation tree if $\soln$ satisfies $C$.
(We make use of analogous lemmas for type well-formedness rules as well.)
\qed
\end{proof}
